MS Technical Paper: \\ Placement Algorithms for Heterogenous FPGAs

### 2025-03-29 latex/paper/paper.tex
more paper work

    Configurable logic devices have undergone significant evolution over the past four decades. 
    The journey began with the Programmable Logic Array (PLA) in the early 1970s. 
    The PLA implemented output logic using a programmable-OR and programmable-AND plane that formed a sum-of-products equation for each output through programmable fuses. 
    Around the same time, the Programmable Array Logic (PAL) was introduced. 
    The PAL simplified the PLA by fixing the OR gates, resulting in a fixed-OR, programmable-AND design, which sacrificed some logic flexibility to simplify its manufacture.

@@

    This brings us to modern day FPGA architectures.
    To meet the needs of increasingly complex designs, FPGA vendors introduced heterogeneous FPGAs. 
    In these devices, hard macros such as Block RAM (BRAM) and Digital Signal Processing (DSP) slices are integrated into the programmable logic fabric along with CLBs. 
    This design enables the direct instantiation of common subsystems like memories and multipliers, without having to implement them solely with CLBs. 
    Major vendors such as Xilinx and Altera now employ heterogeneous island-style architectures in their devices. 
    As designs become increasingly large and complex, FPGAs become increasingly heterogenous as they incorporate a wider variety of hard macros, including embedded CPUs and communications blocks, into the fabric.

    Xilinx introduced the 7-Series architecture in 2010 which followed the heterogeneous island-style architecture. 
    Then in 2013 Xilinx unveiled the Ultrascale architecture which made several improvements on the 7-Series architecture.
    Although the cutting edge of FPGAs follow the Ultrascale architecture, this paper focuses on the former 7-Series devices due to their greater accessibility for beginner users and compatibility with open source tooling.

    Figure~\ref{fig:hierarchy} illustrates the architectural hierarchy of an Ultrascale Xilinx FPGA. 
    The architecture hierarchy is the same for the 7-Series devices. 
    At its most basic level, an FPGA is a vast physical array of replicated atomic components known as Basic Elements of Logic (BELs). 
    These BELs, which include programmable lookup tables (LUTs), flip flops (FFs), and programmable interconnects, form the fundamental building blocks for custom digital circuit implementations on FPGAs.

    In the FPGAs 7-Series FPGAs, additional BELs like Block RAM (BRAM) and Digital Processing Slices (DSPs) are also available, creating a more heterogeneous programmable logic fabric. 
    These dedicated resources allow designers to implement common digital structures, such as multipliers and memories, without having to recreate them from scratch using only LUTs and flip flops.

@@

    Throughout the flow, additional tools, such as constraint managers and timing analyzers, are used to enforce design rules and verify that the implementation meets required performance specifications.

@@

\section{RapidWright API}
    \begin{figure}
        \centering
        \includegraphics[width=6.5cm]{figures/vivado_dcps.png}
        \caption{Showing how RapidWright can take a design checkpoint at different stages of the Vivado toolchain, make customizations, then feed it back into the toolchain.}
        \label{fig:device_carry_chain_routing}
    \end{figure}

    DO WE REALLY NEED THIS SECTION? JUST REFERENCE THE RW DOCUMENTATION.

    The RapidWright API offers a user-friendly toolkit to take pre-implemented designs, that is, designs that have already been placed and/or routed by Vivado, and make custom optimizations to fit complex design criteria. 
    RapidWright has several "BlockPlacers", a general Router, but no general Placer. 
    We use this API to implement a crude general placer, in our case, a simulated annealer. 
    (Show what parts of the FPGA toolchain that RapidWright already has and how our placer fits into that toolchain).

    There are three packages of objects that are most relavent to building our placer: the edif package, design package, and device package.

    \subsection{Device Objects}
    The \textbf{device} package contains the classes that correspond to constructs in the hardware and/or silicon devices. 
    The most prominent class, aptly named Device, represents a specific product family member (like the xc7z020 device of the 7-series family) and provides a reference to all of the logical and routing resources available on the device. 
    There device classes can be separated into two groups: logical and routing. 
    Among the logical classes are the BEL, Site, Tile, and ClockRegion objects discussed before. 
    Among the routing classes are BELPin, SitePin, PIP (programmable interconnect point), Wire, and Node.

@@

    \subsection{EDIF Package}
    In Vivado, all designs post synthesis have a logical netlist that can be exported in the Electronic Design Interchange Format (EDIF) netlist format. 
    The synthesis stage of the FPGA toolchain produces an EDIF file, which is a low-level file containing the technology-mapped netlist. 
    The EDIFNetlist is the top level class that contains the netlist and cell libraries. 
    The EDIFNetlist keeps a reference to the top cell which is wrapped in the EDIFDesign class. 
    It also maintains a top cell instance reference that is generated when the file is loaded.

\section{Placement Implementation}
    (Describe the algorithm, then in context of 7-Series arhictecture). \\
    \subsection{PrePacking}
        In the prepacking stage, we traverse the raw EDIF Netlist to identify recurring cell patterns. In particular, we will specifically look for CARRY chains, DSP cascades, and LUT-FF pairs.
        \subsection{CARRY Chains}
            Each SLICE Site contains one CARRY4 BEL.
            Each CLB Tile contains two SLICE Sites.
            CARRY chains span vertically across multiple SLICE Sites/Tiles. (Show picture)
            \begin{figure}
                \centering
                \includegraphics[width=3.0cm]{figures/carry_chain_routes.png}
                \includegraphics[width=3.0cm]{figures/dsp_cascade_routes.png}
                \caption{A Carry Chain of length 6 spanning 6 Tiles. On the left, the cell-only representation. On the right, cells with routing representation.}
                \label{fig:device_carry_chain_routing}
            \end{figure}
        \subsection{DSP Cascades}
            Each DSP Site contains one DSP BEL.
            Each DSP Tile contains two DSP Sites.
            DSP chains can span vertically across multiple DSP Sites/Tiles.

>>>>

    Here we will briefly outline the evolution of configurable logic architecture as well as the FPGA design flow. 
    However, readers who are already familiar with basic FPGA architecture and design flow may skip to the next section on Xilinx 7-Series Architecture. 
    Configurable logic devices have undergone significant evolution over the past four decades. 
    The journey began with the Programmable Logic Array (PLA) in the early 1970s. 
    The PLA implemented output logic using a programmable-OR and programmable-AND plane that formed a sum-of-products equation for each output through programmable fuses. 
    Around the same time, the Programmable Array Logic (PAL) was introduced. 
    The PAL simplified the PLA by fixing the OR gates, resulting in a fixed-OR, programmable-AND design, which sacrificed some logic flexibility to simplify its manufacture.

@@

    This brings us to modern day FPGA architectures.
    To meet the needs of increasingly complex designs, FPGA vendors introduced heterogeneous FPGAs. 
    In these devices, hard macros such as Block RAM (BRAM) and Digital Signal Processing (DSP) slices are integrated into the programmable logic fabric along with CLBs. 
    This design enables the direct instantiation of common subsystems like memories and multipliers, without having to implement them solely with CLBs. 
    Major vendors such as Xilinx and Altera now employ heterogeneous island-style architectures in their devices. 
    As designs become increasingly large and complex, FPGAs become increasingly heterogenous as they incorporate a wider variety of hard macros into the fabric.

@@

@@

    Xilinx introduced the 7-Series architecture in 2010 which followed the heterogeneous island-style architecture described previously. 
    Then in 2013 Xilinx unveiled Ultrascale architecture which kept the same heterogenous island-style architecture but made several optimization over the 7-Series architecture.
    Although the latest FPGAs follow the Ultrascale architecture, this paper focuses on the 7-Series FPGAs due to their greater accessibility for beginner users and compatibility with open source tooling.

    Figure~\ref{fig:hierarchy} illustrates the architectural hierarchy of an Ultrascale Xilinx FPGA. 
    The architecture hierarchy is the same for the 7-Series devices. 
    The only difference between Ultrascale and 7-Series is simply the design of the CLBs (Sites and BELs) themselves. 
    At its most basic level, an FPGA is a vast physical array of replicated atomic components known as Basic Elements of Logic (BELs). 
    These BELs, which include LUTs, FFs, BRAMs, DSPs, and programmable interconnects, form the fundamental building blocks for custom digital circuit implementations on FPGAs.

\section{RapidWright API}
    \begin{figure}
        \centering
        \includegraphics[width=6.5cm]{figures/vivado_dcps.png}
        \caption{Showing how RapidWright can take a design checkpoint at different stages of the Vivado toolchain, make customizations, then feed it back into the toolchain.}
        \label{fig:vivado_dcps}
    \end{figure}

    We use the RapidWright API to implement a custom placer for Xilinx 7-Series FPGAs.
    RapidWright is an open source Java framework that enables netlist implementation manipulation for modern Xilinx FPGAs. 
    Figure \ref{fig:vivado_dcps} shows how the API can take design checkpoints in the form of .dcp files generated at any stage of the design implementation, allow the user to make custom optimizations, and reintegrate it back into the design flow. 
    This allows the user to substitute any stage in the default FPGA design flow offered by Vivado with user-defined optimizations and implementations to meet complex design requirements. 
    There are three packages of objects that are most relavent to building our placer: the edif package, design package, and device package.

    maybe just list the relevant objects?

    \subsection{EDIF Objects}
    In Vivado, all designs post synthesis have a logical netlist that can be exported as an Electronic Design Interchange Format (EDIF) file or within a .dcp checkpoint file. 
    This is the result of the synthesis tool parsing the raw Verilog source files, inferring logical behavior, and instantiating primitive cells to emulate the desired logical behavior. 
    The final stage is often referred to as "technology mapping", as the primitive cells are specific to the FPGA architecture and consequently the FPGA manufacturer.

    The EDIFNetlist is the top level class that contains the netlist and cell libraries. 
    The EDIFNetlist keeps a reference to the top cell which is wrapped in the EDIFDesign class. 
    It also maintains a top cell instance reference that is generated when the file is loaded.

    \begin{figure*}
        \centering
        \includegraphics[width=\textwidth]{figures/routing_objects.png}
        \caption{Examples of routing objects in a Xilinx FPGA}
        \label{fig:routing_objects}
    \end{figure*}

    \subsection{Device Objects}
    The \textbf{device} package contains the classes that correspond to constructs in the hardware and/or silicon devices. 
    The most prominent class, aptly named Device, represents a specific product family member (like the xc7z020 device of the 7-series family) and provides a reference to all of the logical and routing resources available on the device. 
    The device classes can be separated into two main groups: logical and routing. 
    Among the logical classes are the BEL, Site, Tile, and ClockRegion objects as shown previously in Figure \ref{fig:hierarchy}. 
    Shown in figure \ref{fig:routing_objects} are the routing classes including BELPin, SitePin, PIP (programmable interconnect point), Wire, and Node.

@@

    In its current state, RapidWright has several SA-based "BlockPlacers" that can take pre-implemented macros, a general Router, but no general Placer. 
    To fill this gap, we attempt to build a general Placer that can take a raw EDIF netlist and create a placed design. 
    As mentioned before, our placer will be divided into three substages: prepacking, packing, and placement.
    We use this API to implement a crude general placer, in our case, a simulated annealer. 
    (Show what parts of the FPGA toolchain that RapidWright already has and how our placer fits into that toolchain).

\section{Placement Implementation}
    (Describe the algorithm, then in context of 7-Series arhictecture). \\
    \subsection{PrePacking}
        In the prepacking stage, we traverse the raw EDIF Netlist to identify recurring cell patterns. 
        In particular, we will specifically look for CARRY chains, DSP cascades, and LUT-FF pairs as these patterns are most prevalent in all FPGA designs.
        \subsection{CARRY Chains}
            Each SLICE Site contains one CARRY4 BEL.
            Each CLB Tile contains two SLICE Sites.
            CARRY chains span vertically across multiple SLICE Sites/Tiles. (Show picture)
            \begin{figure}
                \centering
                \includegraphics[width=3.0cm]{figures/carry_chain_routes.png}
                \includegraphics[width=3.0cm]{figures/dsp_cascade_routes.png}
                \caption{A Carry Chain of length 6 spanning 6 Tiles. On the left, the cell-only representation. On the right, cells with routing representation.}
                \label{fig:device_carry_chain_routing}
            \end{figure}
        \subsection{DSP Cascades}
            Each DSP Site contains one DSP BEL.
            Each DSP Tile contains two DSP Sites.
            DSP chains can span vertically across multiple DSP Sites/Tiles.
